Viscous Transport in Eroding Porous Media

### 2019-09-26 docs/transport/erosion_transport.tex
worked through rest of paper and Nick's comments

\begin{abstract} 
  We extend our previous work, [Journal of Computational Physics, {\bf
  375}:1--21, 2018], to simulate dense suspensions of fluid-mechanical
  erosion of two-dimensional erosion of porous media. By using a
  Barycentric quadrature rule, we resolve interactions
  between nearly touching bodies without requiring an unfeasible number
  of discretization points.  Since the Barycentric quadrature rule is
  computationally expensive, it is only used when necessary, and the
  spectrally accurate trapezoid rule is used otherwise. Our final fluid
  solver only requires $\mathcal{O}(N)$ iterations per time step. Next,
  we simulate and analyze transport through eroded porous geometries.
  Our results demonstrate the effects of erosion on tortuosity,
  dispersion, and pore size distributions.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998}, velocity scales can
be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}, \awk{and
the scales of a single porous media can range over several orders of
magnitude.}
%
\todo[inline]{Nick: I am confused by the last part of the sentence (highlighted in purple). What `scale' are you referring to? Length? You already described the length scales in the first part of the sentence. Maybe you mean to distinguish between the grain length scales and the overall length scales of the system. If so, you can use adjectives like `grain' and `macroscopic' or `overall' to clarify that.} 
%
Numerical methods that \st{accurately address all of these} \nick{resolve this range of} scales offer the ability to: (i) characterize dispersion~\cite{saf1959}, (ii) quantify mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and (iii) potentially develop meaningful constitutive relationships that link the microscopic and macroscopic realms~\cite{mil-chr-imh-mcb-ped1998}. 
%
\todo[inline]{Nick: I reworded the above sentence. See if you agree with my changes.}
%
Examples of coarse-grained models for porous media flow include
pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in time in applications involving melting~\cite{bec-vis1998, rycroft2016asymmetric, favier2019rayleigh, morrow2019moving}, dissolution~\cite{kan-zha-che-he2002, mac2015shape, moore2017riemann, wykes2018self}, deposition~\cite{joh-eli1995, hewett2018modelling}, biofilm growth~\cite{tan-val-wer2015}, and \nick{crack formation} \cite{cho2019crack}.  
%
\todo[inline]{Nick: I added some references. Check out the last one on crack formation in shrinkable, granular packings. It is really cool work done by an excellent experimentalist Sujit Datta. I met him and he voiced a strong interest in our work, so wanted to make you aware.}
%
We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, \nick{and industrial} 
applications \cite{berhanu2012shape, hewett2017evolution, lopez2018cfd, allen2019sde, amin2019role}.
\todo[inline]{Nick: Check out the first reference, Berhanu, Kudrolli, Rothman PRE 2012. Maybe we can cite it a couple more times throughout since it has both the words `erosion' and `granular bed' in the title :)}
When a porous geometry erodes, certain qualitative characteristics are unveiled that affect transport through the geometry.
For example, an eroded geometry may contain channels of high porosity, \nick{which, though few in number and modest in volume fraction, transmit a large portion of the flux}. This arrangement results in velocities
that vary over several orders of magnitude~\cite{all-hea-lab-rei2002}.
%
\todo[inline]{Nick: I rearranged the above two sentences. See if you like it.}
%
Moreover, \nick{channelization creates a heterogeneous and anisotropic medium properties}, \st{and
this affects} \nick{affecting} the transport of \st{quantities} \nick{tracers} such as
contaminants~\cite{cve-che-wen1996, dag1987, kon-bre1978} and
heat~\cite{nil-sto1990, ree-sto1995}.  \awk{To characterize transport in
eroded geometries, we complete two undertakings: forming realistic
eroded porous geometries and characterizing transport through these
geometries}.  
\todo[inline]{Nick: The purple sentence is redundant: `To \underline{characterize transport} ... we undertake ... \underline{characterizing transport} through these geometries. Maybe try something like this:}
\nick{This study consists of two main undertakings: First, high-fidelity simulations of porous-medium erosion, and, second, the characterization of transport through the resulting geometries.}
Our modeling efforts build on previous work \cite{ris-moo-chi-she-zha2012, moore2013self, moore2017riemann}, in particular recent numerical methods developed to simulate erosion in the regime of Stokes flow \cite{qua-moo2018}. We, however, make a few key improvements to the numerical methods from \cite{qua-moo2018} to enable simulations of more realistic, dense suspensions of eroding grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To characterize transport in the resulting geometries, \st{we apply statistical methods to develop coarse-grained variables} \nick{examine coarse-grained variables through statistical analysis of tracer trajectories}.
\todo[inline]{I'm not sure about my modification above, but see if it gets across the message that was intended.}

@@

\st{Because of the length and velocity scales in hydrological applications
such as groundwater flow} \nick{Owing to the scales present in groundwater flow} \cite{bea1972}, we model the hydrodynamics with
the two-dimensional incompressible Stokes equations.  \st{To couple erosion
with the fluid} Meanwhile, individual grains erode at a rate \st{that is} proportional to \st{the magnitude of} the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the vorticity and the shear stress are equivalent on the grains in the zero Reynolds number
limit, we also compute the vorticity in the fluid bulk \nick{for visualization purposes} (Figure~\ref{fig:Eroding50vort}).  A variety of numerical methods for
the fluid solver are available including lattice-gas or lattice
Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Transport, Tracers, and Tortuosity}
\label{sec:transport}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Erosion in porous geometries leads to particular phenomena such as
channelization, and we are interested in characterizing transport in
such geometries.  In our previous work~\cite{qua-moo2018}, we examined
the effect of erosion on the area fraction, flow rate, and the total
drag.  However, to characterize transport, other quantities must be
examined.  Here, we compute the anomalous dispersion rate, the
tortuosity, and the distribution of the pore sizes.  The first two
metrics can be defined in terms of streamlines governed by an advection
autonomous equation~\eqref{eqn:tracers}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in a porous media depends on the porosity, it also
depends on the placement of the grains.  In particular, the grain
placement affects velocity scales~\cite{ali-par-wei-bre2017},
correlation structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019}, and more. In addition,
the grain configuration is important for pore network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} that rely on
descriptions of the pore throat sizes. To characterize the grain
placement, we compute distributions of pore sizes between neighboring
grains.  To define neighboring grains, we form a Delaunay triangulation
using nodes placed at the center of each eroding grain and at a collection
of points around the boundary of the porous region.  Then, we say that
two grains are neighbors if their centers share an edge of the
triangulation~\cite{dea-qua-bir-jua2018}.  Finally, we compute the
minimum distance between all pairs of neighboring grains.  In
Figure~\ref{fig:Eroding100gap}(a), we plot a histogram of the pore size
of 100 eroded bodies.  The black curve is an unnormalized Weibull
distribution with the same mean and variance as the measured data.  In
Figure~\ref{fig:Eroding100gap}(b), we plot illustrate the pore throat
sizes by drawing the shortest lines between neighboring grains.  We do
not consider pores between eroding bodies and the solid wall $\Gamma$,
so some of the grains near the porous region boundary only have two
neighbors.

@@

The erosion rate has a loss of regularity at stagnation points, and this
inevitably leads to corners developing on the eroding bodies.
Therefore, as described in Section~\ref{sec:formulation} and our
previous work~\cite{qua-moo2018}, we introduce a curvature penalization
term and a Gaussian smoothing step.  For all examples, we use the
smoothing parameters $\eps=15/256$ and $\sigma=10/256$.

@@

We consider an eroding body that is initially a circle of radius
$r=0.4$.  We discretize the eroding body with $N_\iin=256$
discretization points, and its center is chosen so that the distance
between the body and the solid wall is $h/2$, $h/4$, and $h/10$, where
$h = 2\pi r/N_\iin$.  If we used the trapezoid rule and required an
error that is comparable to the Barycentric quadrature rule, the body
with an initial distance of $h/10$ from the solid wall would require
$6,400$ discretization points, and the outer wall would require $50,000$
discretization points.  For all three simulations, we use the time step
size $\Delta t=10^{-4}$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider 20 eroding bodies that are discretized with $N_\iin=256$
points and a time step size of $\Delta t = 10^{-4}$.  The background
flow is the Hagen-Poiseuille flow
\begin{align}
  \UU(\xx)=U \left[
  \begin{array}{c}
    1-y^2 \\ 0
  \end{array}
  \right],
\end{align}
where the flow rate $U$ is chosen so that the pressure drop from $x=-2$
to $x=2$ is held fixed at 2.  Snapshots of the bodies at four equispaced
times are shown in Figure~\ref{fig:Eroding20vort}.  The color is the
vorticity, and the erosion rate is proportional to the vorticity.
Initially, several of the eroding bodies are closer to the outer wall
than the $5h$ threshold required to guarantee that the trapezoid rule
achieves machine precision.  In particular, the distance between bodies
1, 6, 13, and 15 and the outer wall is $1.3h$, $2.9h$, $2.8h$, and
$1.3h$, respectively, where $h$ is the arclength spacing of the outer
wall $\Gamma$.  In addition, the distance between several pairs of
eroding bodies, including 1 \& 6, 3 \& 9, 6 \& 8, and 14\& 18, is too
small to be accurately resolved with the trapezoid rule.  By using the
Barycentric quadrature rule, the interaction between these
nearly-touching bodies can be accurately resolved at the prescribed
resolution, and erosion can be simulated until all the bodies have
vanished.

@@

Next, we use the streamlines to compute the local tortuosity and the
anomalous dispersion rate of eroding geometries.  To achieve statistical
convergence, we use $N_p = 1000$ streamlines~\cite{bel-sal-rin1992}. To
compute the tortuosity, we require the velocity at the inlet $x=-1$.
These normalized velocities are plotted in
Figure~\ref{fig:Eroding20tort}(a) for the eroded geometry with porosity
62.9\% (see Figure~\ref{fig:Eroding20tort}(c)).  The velocities are
similar to the work of Matyka et al.~\cite{mat-kha-koz2008} (see Figure
4(a)), except that our cross-section, by construction, does not cut
through any of the grains.  Next, in Figure~\ref{fig:Eroding20tort}(b),
we plot the local tortuosity~\eqref{eqn:localTort} by calculating the
relative length of each streamline as it traverses the channel from
$x=-1$ to $x=1$.  The local tortuosity ranges from 1 to 1.27, meaning
that one of the streamlines is 27\% longer than it would have been if
the grains were absent.  The average streamline length is 9.79\% longer
than if the grains had been absent, so the tortuosity of the eroded
channel is $1.098$.  Again, comparing the local tortuosity to
Matyka~\cite{mat-kha-koz2008} (Figure 4(b)), the results are
qualitatively similar. However, since our initial cross-section does not
cut through the grains, the local tortuosity does not have any gaps.
The discontinuities in the local tortuosity happens when nearby
streamlines diverge on opposite sides of a grain.  In
Figure~\ref{fig:Eroding20tort}(c), we plot the pairs of streamlines that
correspond to the ten largest jumps in the local tortuosity.
Corresponding pairs are plotted in the same color.

@@

\begin{figure}[H]
\begin{center}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort1}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort150}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort300}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort450}
\end{center}
\caption{\label{fig:ErodingLow20vort} The erosion of 20 nearly touching
bodies, each discretized with $N_\iin = 256$ points, in a
Hagen-Poiseuille flow. The 4 snapshots are evenly spaced in time, and
the color is the fluid vorticity. In addition to the channels that
develop between the bodies and the solid walls, erosion leads to two
main channels through the geometry---one in the top half and one near
the middle.}
\end{figure}

@@

\todo[inline]{HERE}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{100 eroding bodies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
As a final example, we consider 100 eroding bodies with an initial
porosity near 50\%.  Snapshots of the configurations and vorticity are in
Figure~\ref{fig:Eroding100vort}.  Again, by resolving near-contacts with
the Barycentric quadrature rule, a stable simulation is possible with
only $N_\iin = 256$ discretization points.  We use a time step size of
$\Delta t = 10^{-4}.$

\begin{figure}[H]
 \begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =0.9\linewidth]{./figs/100b_50}
\caption{100 bodies, $\phi = 55.54\%$}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =1.1\linewidth]{./figs/100b_100}
\caption{99 bodies, $\phi = 62.98\%$}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =0.9\linewidth]{./figs/100b_150}
\caption{94 bodies, $\phi = 72.22\%$}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =1.1\linewidth]{./figs/100b_200}
\caption{82 bodies, $\phi = 83.67\%$}
\end{subfigure}
\caption{\label{fig:Eroding100vort} Erosion of 100 bodies with the
  fixed-pressure-drop condition. The color is the vorticity of fluid. In
  this simulation, we use $N_\iin = 256$ discretization points on the
  body, $N_\out = 1024$ points on the outer wall, a time-step of
  $\Dt=10^{-4}$.  Because of the large number of bodies, erosion creates
  many channels connecting the inlet to the outlet.}
\end{figure}

>>>>

\begin{abstract} 
  We extend upon our previous work, [Journal of Computational Physics,
  {\bf 375}:1--21, 2018], to simulate dense suspensions of
  fluid-mechanical erosion of two-dimensional erosion of porous media.
  The method combines a boundary integral equation (BIE) fluid solver
  and a numerically stable time stepping method for the evolution of the
  eroding bodies. Originally, the eroding bodies had to be sufficiently
  separated to guarantee that the trapezoid rule provided a sufficient
  level of accuracy. In this work, we develop resolve interactions
  between nearly touching bodies with a Barycentric quadrature rule
  whose error is independent of the geometry.  Since the Barycentric
  quadrature rule is computationally expensive, it is only used when
  necessary, and the spectrally accurate trapezoid rule is used
  otherwise. Our final fluid solver only requires $\mathcal{O}(N)$
  iterations per time step. Next, we simulate and analyze transport
  through eroded porous geometries.  Our results demonstrate the effects
  of erosion on tortuosity, dispersion, and pore size distributions.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\cite{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\cite{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\cite{saf1959}, (ii) quantify
mixing~\cite{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) potentially develop meaningful constitutive relationships that
link the microscopic and macroscopic
realms~\cite{mil-chr-imh-mcb-ped1998}.  Examples of coarse-grained
models for porous media flow include pressure-saturation-permeability
relationships~\cite{mil-chr-imh-mcb-ped1998}, permeability-porosity
relationships~\cite{dar-mcc1998, car1937, won-kop-tom1984}, first
passage time distributions~\cite{ber-sch-sil2000, hym-den-hag-kan2019,
cve-che-wen1996}, tortuosities~\cite{hak-com-den2019, mat-kha-koz2008,
dud-koz-mat2011, kop-kat-tim1996}, concentration
distributions~\cite{ica-den2019, bel-sal-rin1992}, flow
distributions~\cite{ali-par-wei-bre2017}, geometry
connectivity~\cite{knu-car2005, wes-blo-gra2001}, pore
distributions~\cite{ali-par-wei-bre2017}, anomalous
dispersion~\cite{dea-qua-bir-jua2018, den-cor-sch-ber2004,
sie-ili-pri-riv-gua2019}, and more.

Flow in porous media is further complicated when there is a two-way
coupling between the flow and the geometry. For example, the grains of a
porous media evolve in time in applications involving
melting~\cite{bec-vis1998, rycroft2016asymmetric, favier2019rayleigh,
morrow2019moving}, dissolution~\cite{kan-zha-che-he2002, mac2015shape,
moo2017, wykes2018self}, deposition~\cite{joh-eli1995,
hewett2018modelling}, biofilm growth~\cite{tan-val-wer2015}, and crack
formation~\cite{cho2019crack}. We focus on erosion, a fluid-mechanical
process that is prevalent in many geophysical, hydrological, and
industrial applications~\cite{berhanu2012shape, hewett2017evolution,
lopez2018cfd, allen2019sde, amin2019role}.  When a porous geometry
erodes, certain qualitative characteristics are unveiled that affect
transport through the geometry.  For example, an eroded geometry may
contain channels of high porosity, which, though few in number and
modest in volume fraction, transmit a large portion of the flux. This
arrangement results in velocities that vary over several orders of
magnitude~\cite{all-hea-lab-rei2002}.  Moreover, channelization creates
a heterogeneous and anisotropic medium properties affecting the
transport of tracers such as contaminants~\cite{cve-che-wen1996,
dag1987, kon-bre1978} and heat~\cite{nil-sto1990, ree-sto1995}.

This study consists of two main undertakings: first, high-fidelity
simulations of eroding porous media, and, second, characterizing
transport through the resulting geometries. Our modeling efforts build
on previous work~\cite{ris-moo-chi-she-zha2012, moore2013self, moo2017},
in particular recent numerical methods developed to simulate erosion in
the regime of Stokes flow~\cite{qua-moo2018}. We, however, make a few
key improvements to the numerical methods from~\cite{qua-moo2018} to
enable simulations of more realistic, dense suspensions of eroding
grains (Figures~\ref{fig:Eroding50vel} and~\ref{fig:Eroding50vort}).  To
characterize transport in the resulting geometries, examine
coarse-grained variables through statistical analysis of tracer
trajectories.

@@

Owing to the scales present in groundwater flow~\cite{bea1972}, we model
the hydrodynamics with the two-dimensional incompressible Stokes
equations.  Meanwhile, individual grains erode at a rate proportional to
the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, moore2013self, par-izu2000}.  Since the
vorticity and the shear stress are equivalent on the grains in the zero
Reynolds number limit, we also compute the vorticity in the fluid bulk
for visualization purposes (Figure~\ref{fig:Eroding50vort}).  A variety
of numerical methods for the fluid solver are available including
lattice-gas or lattice Boltzmann~\cite{kop-kat-tim1996, dar-mcc1998},
finite volumes~\cite{suo-liu-gan2019, den-ica-hid2018,
sie-ili-pri-riv-gua2019}, finite differences~\cite{leb-ded-dav-bou2007,
knu-car2005}, and cellular automata~\cite{rot1988}.  However, simulating
erosion requires a method that is high-order, can resolve
nearly-touching arbitrarily shaped grains, and can accurately compute
derivatives of the velocity field.  Since the governing equations are
linear and homogeneous, we address these challenges by converting the
fluid equations to a boundary integral equation (BIE).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Transport, Tracers, and Tortuosity}
\label{sec:transport}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Erosion in porous geometries leads to particular phenomena such as
channelization~\cite{berhanu2012shape}, and we are interested in
characterizing transport in such geometries.  In our previous
work~\cite{qua-moo2018}, we examined the effect of erosion on the area
fraction, flow rate, and the total drag.  However, to characterize
transport, other quantities must be examined.  Here, we compute the
anomalous dispersion rate, the tortuosity, and the distribution of the
pore sizes.  The first two metrics can be defined in terms of
streamlines governed by an advection autonomous
equation~\eqref{eqn:tracers}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While transport in a porous media depends on the porosity, it also
depends on the placement of the grains.  In particular, the grain
placement affects velocity scales~\cite{ali-par-wei-bre2017},
correlation structures~\cite{leb-ded-dav-bou2007}, contaminant
transport~\cite{knu-car2005},
channelization~\cite{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
more. In addition, the grain configuration is important for pore network
models~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} that rely on
descriptions of the pore throat sizes. To characterize the grain
placement, we compute distributions of pore sizes between neighboring
grains.  To define neighboring grains, we form a Delaunay triangulation
using nodes placed at the center of each eroding grain and at a
collection of points around the boundary of the porous region.  Then, we
say that two grains are neighbors if their centers share an edge of the
triangulation~\cite{dea-qua-bir-jua2018}.  Finally, we compute the
minimum distance between all pairs of neighboring grains.  In
Figure~\ref{fig:Eroding100gap}(a), we plot a histogram of the pore size
of 100 eroded bodies.  The black curve is an unnormalized Weibull
distribution with the same mean and variance as the measured data.  In
Figure~\ref{fig:Eroding100gap}(b), we plot illustrate the pore throat
sizes by drawing the shortest lines between neighboring grains.  We do
not consider pores between eroding bodies and the solid wall $\Gamma$,
so some of the grains near the porous region boundary only have two
neighbors.  To analyze the effect of erosion on pore sizes, we use a
single Delaunay triangulation to define the neighbors up until a body
has completely eroded. At this point, a new Delanuay triangulation is
formed.

@@

The erosion rate has a loss of regularity at stagnation points, and this
inevitably leads to corners developing on the eroding bodies.
Therefore, as described in Section~\ref{sec:formulation} and our
previous work~\cite{qua-moo2018}, we introduce a curvature penalization
term and a Gaussian smoothing step.  For all examples, we use the
smoothing parameters $\eps=15/256$ and $\sigma=10/256$, and the time
step size is $\Delta t = 10^{-4}$.

@@

We consider an eroding body that is initially a circle of radius
$r=0.4$.  We discretize the eroding body with $N_\iin=256$
discretization points, and its center is chosen so that the distance
between the body and the solid wall is $h/2$, $h/4$, and $h/10$, where
$h = 2\pi r/N_\iin$.  If we used the trapezoid rule and required an
error that is comparable to the Barycentric quadrature rule, the body
with an initial distance of $h/10$ from the solid wall would require
$6,400$ discretization points, and the outer wall would require $50,000$
discretization points.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Medium Porosity}
\label{sec:Eroding20}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider 20 eroding bodies that are discretized with $N_\iin=256$
points.  The background flow is the Hagen-Poiseuille flow
\begin{align}
  \UU(\xx)=U \left[
  \begin{array}{c}
    1-y^2 \\ 0
  \end{array}
  \right],
\end{align}
where the flow rate $U$ is chosen so that the pressure drop from $x=-2$
to $x=2$ is held fixed at 2.  Snapshots of the bodies at four equispaced
times are shown in Figure~\ref{fig:Eroding20vort}.  The color is the
vorticity, and the erosion rate is proportional to the vorticity.
Initially, several of the eroding bodies are closer to the outer wall
than the $5h$ threshold required to guarantee that the trapezoid rule
achieves machine precision.  In particular, the distance between bodies
1, 6, 13, and 15 and the outer wall is $1.3h$, $2.9h$, $2.8h$, and
$1.3h$, respectively, where $h$ is the arclength spacing of the outer
wall $\Gamma$.  In addition, the distance between several pairs of
eroding bodies, including 1 \& 6, 3 \& 9, 6 \& 8, and 14\& 18, is too
small to be accurately resolved with the trapezoid rule.  By using the
Barycentric quadrature rule, the interaction between these
nearly-touching bodies can be accurately resolved at the prescribed
resolution, and erosion can be simulated until all the bodies have
vanished.

@@

Next, we use the streamlines to compute the tortuosity and the anomalous
dispersion rate of eroding geometries.  To achieve statistical
convergence, we use $N_p = 1000$ streamlines~\cite{bel-sal-rin1992}. To
compute the tortuosity, we require the velocity at the inlet $x=-1$.
These normalized velocities are plotted in
Figure~\ref{fig:Eroding20tort}(a) for the eroded geometry at porosity
$\phi = 62.9\%$ (Figure~\ref{fig:Eroding20tort}(c)).  The velocities are
similar to the work of Matyka et al.~\cite{mat-kha-koz2008} (see Figure
4(a)), except that our cross-section, by construction, does not cut
through any of the grains.  Next, in Figure~\ref{fig:Eroding20tort}(b),
we plot the local tortuosity~\eqref{eqn:localTort} by calculating the
relative length of each streamline as it traverses the channel from
$x=-1$ to $x=1$.  The local tortuosity ranges from 1 to 1.27, meaning
that one of the streamlines is 27\% longer than it would have been if
the grains were absent.  The average streamline length is 9.79\% longer
than if the grains had been absent, so the tortuosity of the eroded
channel is $1.098$.  Again, comparing the local tortuosity to
Matyka~\cite{mat-kha-koz2008} (Figure 4(b)), the results are
qualitatively similar. However, since our initial cross-section does not
cut through the grains, the local tortuosity does not have any gaps.
The discontinuities in the local tortuosity happens when nearby
streamlines diverge on opposite sides of a grain.  In
Figure~\ref{fig:Eroding20tort}(c), we plot the pairs of streamlines that
correspond to the ten largest jumps in the local tortuosity.
Corresponding pairs are plotted in the same color.

@@

\begin{figure}[H]
\begin{center}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort1}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort150}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort300}
  \includegraphics[height=0.225\textwidth]{./figs/20b2_vort450}
\end{center}
\caption{\label{fig:ErodingLow20vort} The erosion of 20 nearly touching
bodies, each discretized with $N_\iin = 256$ points, in a
Hagen-Poiseuille flow. The four snapshots are evenly spaced in time, and
the color is the fluid vorticity. In addition to the channels that
develop between the bodies and the solid walls, erosion leads to two
main channels through the geometry---one in the top half and one near
the middle.}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{100 eroding bodies}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
As a final example, we consider 100 eroding bodies with an initial
porosity near 50\%.  Snapshots of the configurations and vorticity are
in Figure~\ref{fig:Eroding100vort}.  Again, by resolving near-contacts
with the Barycentric quadrature rule, a stable simulation is possible
with only $N_\iin = 256$ discretization points.

\begin{figure}[H]
 \begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =0.9\linewidth]{./figs/100b_50}
\caption{100 bodies, $\phi = 55.54\%$}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =1.1\linewidth]{./figs/100b_100}
\caption{99 bodies, $\phi = 62.98\%$}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =0.9\linewidth]{./figs/100b_150}
\caption{94 bodies, $\phi = 72.22\%$}
\end{subfigure}%
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[width =1.1\linewidth]{./figs/100b_200}
\caption{82 bodies, $\phi = 83.67\%$}
\end{subfigure}
\caption{\label{fig:Eroding100vort} The erosion of 100 nearly
touching grains, each discretized with $N_\iin = 256$ points, in a
Hagen-Poiseuille flow. The color is the vorticity of fluid. The four
snapshots are evenly spaced in time, and the color is the fluid
vorticity.  Because of the large number of bodies, erosion creates many
channels connecting the inlet to the outlet.}
\end{figure}

We compute the tortuosity using both the
Lagrangian~\eqref{eqn:tortuosity1} and Eulerian
methods~\eqref{eqn:tortuosity2}. Therefore, we must compute the velocity
at $N_p = 1000$ points along the inlet $x=-1$. These normalized
velocities are plotted in Figure~\ref{fig:Eroding100tort}(a) for the
eroded geometry at porosity $\phi = 63.0\%$
(Figure~\ref{fig:Eroding100tort}(c)).  The initial velocity of the
tracers is qualitatively similar to the 20 body example
(Figure~\ref{fig:Eroding20tort}(a)), except with additional maximum and
minimums because of the additional bodies.  Next, in
Figure~\ref{fig:Eroding100tort}(b), we plot the local tortuosity by
finding the length of the streamlines as they pass from $x=-1$ to $x=1$.
Compared to Figure~\ref{fig:Eroding20tort}(b), the local tortuosity is
much more discontinuous.  These discontinuities can be explained by
examining the trajectories of tracers in
Figure~\ref{fig:Eroding100tort}(c).  Here, there are many instances of
nearby streamlines that are deflected apart from one another as they
tend to a stagnation point in the flow, and this results in trajectories
with significantly different lengths.  At this porosity, one of the
tracers travels 25.5\% farther than it would have if the bodies had been
absent, and the average tracer travelled 12\% farther than it would have
if the bodies were absent, so the tortuosity is $1.12$.
